---
title: "UMCCR bcbio single patient config"
author: "Oliver Hofmann"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    toc: false
    toc_float: false
    code_folding: hide
---

```{r custom, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyr)
library(rmarkdown)
library(dplyr)
library(DT)
library(readr)
library(forcats)
library(stringr)
library(janitor)
library(googledrive)
library(aws.s3)
library(aws.lambda)
library(here)
library(skimr)
library(purrr)
```

## Introduction

Code snippets to generate file summaries (for rsync'ing around) and bcbio sample summaries based on a [Google Spreadsheet](https://docs.google.com/spreadsheets/d/1DwvyfVrgr5TIcYtGVXZeIWWRbld-nuX-4o4z2kZHNWs/edit#gid=0), aka the dreadful LIMS stand-in. This version differs from the `bcbioSetup.Rmd` in that it is meant for _single_ patient samples only, i.e., samples are only batched for individual patients not across a whole project. This should be the default approach.

### 1. Setting up project information

This currently requires specifying a 'secondary use' field in the LIMS. Set the "secondary" analysis flag in the code snippet below to match the `secondary use` LIMS field used to highlight samples that need to be processed, regardless of processing status. Any sample with the matching `SECONDARY` entry in the `secondary analysis` column of the spreadsheet will be added as long as it is of the right `TYPE` (WGS for this script). 

The `googledrive` framework requires authentication with oAuth. This can be done interactive, but storing a token simplifies the process; see the [googlesheet authentication vignette](https://rawgit.com/jennybc/googlesheets/master/vignettes/managing-auth-tokens.html) for details. 

```{r project}
SECONDARY <- 'gds'

# Do not change
TYPE <- 'WGS'
PROJECT <- 'All'
```

### 2. Import data from Google spreadsheet

This step generates a data frame from the Google spreadsheet, unifying variable names and simplifying subject IDs for bcbio along the way. It replaces empty `results` cells (which indicate samples that still need processing) with a `-` to distinguish from true NAs and gets rid of whitespace in subject identifiers (although ideally there should not be any empty cells in the sheet in the first place). Otherwise standard data cleanup with `janitor`. We are also creating a timestamped backup of the Google Doc each time it is accessed, just in case.

```{r importData, message=FALSE}
# Create a backup copy each time we access the sample info
tm <- as.POSIXlt(Sys.time(), "UTC", "%Y-%m-%dT%H:%M:%S")
timestamp <- strftime(tm , "%Y-%m-%dT%H%M")
filename <- paste(timestamp, 'backup.csv', sep = '_')
backup_dir <- file.path(here::here('output', 'backup'))
dir.create(backup_dir, recursive = TRUE) # mkdir -p (warns if it already exists):

# Google Drive implementation
lims_key <- drive_find('^Google LIMS$', team_drive = 'LIMS')$id
drive_download(as_id(lims_key), path = file.path(backup_dir, filename), overwrite = TRUE)

# Import downloaded spreadsheet (sheet #1) as a tibble
samples_gs <- read_csv(file.path(backup_dir, filename), col_types = cols(.default = "c", Run = "d"))

# Tweak for analysis
#
# Currently mapping external_subject_id to subject_id (and same for sample id)
#
samples <- samples_gs %>%
  clean_names() %>%
  mutate(secondary_analysis = ifelse(is.na(secondary_analysis), '-', secondary_analysis)) %>%
  remove_empty(c('rows', 'cols')) %>%
  mutate(results = ifelse(is.na(results), '-', results)) %>%
  select(-matches('number_fastqs')) # Drop FASTQ count introduced with new version for now

# Add unique id to each row
samples <- samples %>%
  mutate(row_id = rownames(samples))
```

### 3. Find files ready for processing

Find all samples that either are part of the listed project and still need to be processed, or for which the `secondary flag` matches. Convert different sample naming conventions used throughout into a unified representation.

**Note:** This script relies on internal identifiers (a tuple of `subject_id`, `sample_id` and `library_id`). Where those are missing the script tries to replace them with the equivalent external identifier, or `missing` in the case of `library_id`. This works except for top-ups or where tumor/normal pairings do not share the same internal identifiers. Please make sure older samples have the appropriate internal identifiers set in the LIMS for these use cases. 

```{r subsetSamples}
# Keep rows matching project and type requested; extract path to FASTQ
# and generate new file name for these
bcbio <- samples %>%
  mutate(project_name = if_else(project_owner == '-', 
                                project_name,
                                paste(project_owner, project_name, sep='-'))) %>%
  filter(type == TYPE &
           ((project_name == PROJECT & results == '-') | 
           (secondary_analysis == SECONDARY))) %>% 
  select(illumina_id, fastq, run, project_name, 
         subject_id, sample_id, library_id, 
         external_subject_id, external_sample_id, sample_name, 
         assay, phenotype, source, row_id) %>%
  mutate(subject_id = if_else(subject_id == '-', external_subject_id, subject_id)) %>%
  mutate(sample_id = if_else(sample_id == '-', external_sample_id, sample_id)) %>%
  mutate(library_id = if_else(library_id == '-', 'Missing', library_id)) %>%
  mutate(runname = illumina_id) %>%
#  separate(illumina_id, c(NA, 'machine', NA, NA), sep='_') 
  separate(illumina_id, c('tmpA', 'machine', 'tmpB', 'tmpC'), sep='_') %>%
  select(-tmpA, -tmpB, -tmpC)

# Add a 'target' name: what FASTQs are to be called after retrieval from S3
bcbio <- bcbio %>%
  mutate(targetname = paste(runname, subject_id, sample_id, library_id, sep='_')) %>%
  mutate(targetname = str_replace_all(targetname, '-', '_')) 

# Adding a sample 'description' column used to name the sample within bcbio.
# The description is now a tuple of (subject, sample, library) which should ensure topups
# are handled properly. 
#
# Also look up gds key
bcbio <- bcbio %>%
  mutate(description = paste(subject_id, sample_id, library_id, sep='_')) %>%
  mutate(description = str_replace(description, '_topup.', '')) %>%
  mutate(description = str_replace(description, '_topup', '')) %>%
  mutate(gds_key = paste(sample_id, library_id, sep='_'))
```


### 4. Group samples

Assigning patient/family identifiers for Peddy, assigning batches based on the subject id, and pairing tumor/normals as needed. Samples with the exact same name (but from different runs) are expected to be top-ups and will subsequently be merged using `bcbio_prepare_samples.py`. Finally, remove samples for which we do not have a matched normal.

**Note:** This will likely fail for patients where we have multiple normals that are _not_ top-ups. Might have to handle these cases manually or pick one representative normal.

```{r assignBatches}
# Handle cases where a normal matches more than one tumor sample.
# Courtesy of [Peter Diakumis](https://github.com/pdiakumis)
get_new_batch <- function(batch, phenotype) {
  stopifnot(length(batch) == length(phenotype))
  n <- length(batch)

  new_batch <- vector(mode = "character", length = n)
  x <- table(phenotype)

  if (n > 2) {
    tum_ns <- seq_len(x["tumor"])
    new_batch[phenotype == "tumor"] <- paste0(batch[phenotype == "tumor"], '_', tum_ns)
    # assume only one normal for now
    new_batch[phenotype == "normal"] <- paste0(batch[phenotype == "tumor"], '_', tum_ns, collapse = ";")
  } else {
    new_batch <- batch
  }
  return(new_batch)
}

# Generating the bcbio YAML template. Again, after run 107 use the SampleName instead
# of SampleID to label things. Keep the required columns and add information for peddy
#
# Remove top-up samples (samples with the exact same description) before calculating
# batch assignments. We'll add them back in later. This assumes topups are consistently
# flagged with a `_topup` suffix
#
template <- bcbio %>% 
  mutate(family_id = subject_id) %>%
  mutate(individual_id = subject_id) %>%
  select(samplename = targetname,
         description,
         batch = subject_id,
         source,
         phenotype, family_id, individual_id, row_id,
         external_subject_id,
         external_sample_id,
         external_sample_name = sample_name,
         gds_key,
         project_name,
         )

# Only keep one sample row for samples that _do_ have topups
unique_samples <- template %>%
  select(-row_id) %>%
  group_by(description) %>%
  filter(row_number(description) == 1)

# Identify batches for which we have both tumor and tumor samples
# Well. Technically just testing that we have two distinct phenotypes.
# Could be made more specific (e.g., by ensuring that 'normal' is
# present).
complete_subjects <- template %>%
  group_by(individual_id, batch) %>%
  summarise(phenotype_count = n_distinct(phenotype)) %>%
  filter(phenotype_count >= 2) %>%
  select(individual_id)

complete_subjects <- as.vector(complete_subjects$individual_id)

# Create batch information
batches <- unique_samples %>%
  group_by(batch) %>%
  mutate(new_batch = get_new_batch(batch, phenotype)) 

# Bring back the batch information to the main template
template <- template %>%
  left_join(batches, by = c('description')) %>%
  select(samplename = samplename.x,
         description,
         batch = new_batch,
         phenotype = phenotype.x,
         family_id = family_id.x,
         individual_id = individual_id.x,
         source = source.x,
         external_subject_id = external_subject_id.x,
         external_sample_id = external_sample_id.x,
         external_sample_name = external_sample_name.x,
         gds_key = gds_key.x,
         project_name = project_name.x,
         row_id) 

# Useful when we need non-paired samples
# complete_subjects <- unique(as.vector(template$individual_id))

# Only retain samples that fulfill the 'complete batch' requirement
template <- template %>%
  filter(individual_id %in% complete_subjects)
```

### 5. Generating bcbio csv templates on a per-subject basis:

Generate a file with pointers to the sample FASTQs and their preferred new names as well as a sample descriptor (`.csv`) for bcbio to use as part of it's templating approach. The resulting directory structure can be copied to Spartan, e.g.:

> `scp -r BATCHFILES spartan:/data/cephfs/punim0010/data/Transfer/raijin/`

```{r getGDSPath, eval=TRUE}
# Required information is stored in data portal; set up access credentials
profile <- 'prod'
aws.signature::use_credentials(profile = profile)


get_GDS_Path <- function(illumina_id) {
  
  gds <- paste('gds://umccr-fastq-data-prod/', illumina_id, sep='')
  results <- aws.lambda::invoke_function("data-portal-api-prod-fastq", 
                                         region='ap-southeast-2',
                                         payload=list(gds_path=gds))

  
  results <- results %>%
    as_tibble() %>%
    mutate(samplename = names(fastq_map)) %>%
    unnest_wider(fastq_map) %>% 
    unnest_longer(fastq_list) %>%
    select(samplename, fastq_list) %>%
    rename(gds_key=samplename)

  return(results) 
}

# Get a list of Illumina RunIDs for the current samples
illumina_ids <- unique(bcbio$runname)

# Find all FASTQs for these runs
gds_info <- bind_rows(lapply(illumina_ids, get_GDS_Path))
```

```{r perBatch, eval=TRUE}
write_subject <- function(complete, summary) {

  # Create a unique project name
  subject_name <- paste0(timestamp, '_', PROJECT, '_', TYPE, '_', complete)

  # Set up the project directory
  dir.create(here::here('output', subject_name, 'data'), showWarnings=FALSE, recursive=TRUE)

  # Subset to one individual
  summary <- summary %>%
    filter(individual_id == complete)

  # Find samples in GDS summary
  gds_subset <- gds_info %>%
    filter(gds_key %in% summary$gds_key)
  
  # Find corresponding entries in the Google-Lims
  bcbio_subset <- bcbio %>%
    filter(row_id %in% summary$row_id)

  # Create a file list ready for linking
  gds_link <- bcbio_subset %>%
    right_join(gds_subset, by='gds_key') %>%
    rename(from = fastq_list) %>%
    mutate(to=paste(targetname, 
                     str_replace(basename(from), gds_key, ''),
                     sep='')) %>%
    select(from, to, row_id)
  
  # PBS templating
  file_handle <-  file(here::here('output', subject_name, 'data', 
                                  paste0(subject_name, '_files.sh')))
  write_lines(c('#!/bin/bash',
                '#PBS -P gx8',
                '#PBS -q copyq',
                '#PBS -l walltime=10:00:00',
                '#PBS -l mem=4GB',
                '#PBS -l ncpus=1',
                '#PBS -l software=bcbio',
                '#PBS -l wd',
                '#PBS -l storage=gdata/gx8',
                '',
                '# Adjust path to include IAP',
                'export PATH=/g/data/gx8/local/production/bin:/g/data3/gx8/local/production/bcbio/anaconda/bin:/opt/bin:/bin:/usr/bin:/opt/pbs/default/bin:/g/data/gx8/extras/iap/'),
              file_handle)

  # IAP download requires files to be present - `touch` will do
  write.table(paste('touch', gds_link$to),
              file = here::here('output', subject_name, 'data', 
                                paste0(subject_name, '_files.sh')),
              col.names = FALSE,
              row.names = FALSE,
              quote = FALSE,
              append = TRUE)

  # The actual file download
  write.table(paste('iap files download', gds_link$from, gds_link$to),
              file = here::here('output', subject_name, 'data', paste0(subject_name, '_files.sh')),
              col.names = FALSE,
              row.names = FALSE,
              quote = FALSE,
              append = TRUE)
  
  # Adjust bcbio sample summary with new samplename (filename) information
  summary <- summary %>%
    right_join(gds_link, by='row_id') %>%
    select(-samplename, -from) %>%
    rename(samplename=to) %>%
    relocate(samplename) 

  write.table(summary,
               file = here::here('output', subject_name, 'data', paste0(subject_name, '.csv')),
               row.names = FALSE,
               col.names = TRUE,
               quote = FALSE,
               sep = ',')

    # Add project information for re-use on Gadi
  project_name <- unique(summary$project_name)
  write_file(project_name,
             here::here('output', subject_name, 'data', 'project_name.txt'))
  
}

for (subject in complete_subjects) {
  write_subject(subject, template)
}
```




