---
title: "UMCCR LIMS Sanity Checks"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    theme: paper
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(knitr)
library(tidyr)
library(rmarkdown)
library(dplyr)
library(tidyverse)
library(DT)
library(readr)
library(forcats)
library(stringr)
library(janitor)
library(googlesheets)
library(parsedate)
```

```{r dataPrep}
# Register UMCCR spreadsheet. Use cached authentication
gs_auth(token="./googlesheets_token.rds")
samples_gs <- gs_key('1DwvyfVrgr5TIcYtGVXZeIWWRbld-nuX-4o4z2kZHNWs')

# Tweak for analysis
samples <- samples_gs %>%
  gs_read(ws='Sheet1') %>%
  clean_names() %>%
  remove_empty(c('rows')) %>%
  mutate(subject_id = gsub(' ', '.', subject_id)) %>%
  mutate(results = ifelse(is.na(results), '-', results)) %>%
  filter(project != 'Positive-control') %>%
  filter(project != 'Negative-control')
```

## Manual check

Generating summaries of project names and sample types to detect outliers or typos:

```{r namecheck}
unique(samples$project)
unique(samples$type)

sample_df <- samples %>%
  select(project, type) %>%
  group_by(project, type) %>%
  tally() %>%
  spread(type, n, fill=0) %>%
  as.data.frame() %>%
  column_to_rownames('project')

sample_df %>%
  datatable(rownames=TRUE, 
            options = list(
              pageLength=25,
              dom = 'ft',
              lengthMenu = c(25, 50, 100))) %>%
  formatStyle(names(sample_df),
              background = styleColorBar(range(sample_df), 'lightblue'),
              backgroundSize = '100% 90%',
              backgroundRepeat = 'no-repeat',
              backgroundPosition = 'center')
```

## Storage keys

We recently changed how bcbio's output is stored on Spartan (and eventually on S3). The folder hierarchy should follow a `/data/Results/project_name/timestamp` template, and within that folder you should be able to find the `config`, `final` and `umccrised` folders. Not checking for the latter yet, but checking the Google LIMS to see if the the existing `results` column follows this approach or not.

```{r checkResults}
results_df <- samples %>%
  select(project, results) %>%
  filter(!results %in% c('-', 'n/a', 'Failed'))
```

First pass. Highlight results stored _outside_ of the `Results` folder.

```{r checkFolder}
results_df %>%
  filter(!str_detect(results, '/data/cephfs/punim0010/data/Results')) %>%
  kable()
```

Second pass. Of results stored in the correct `Results` folder flag those not ending with a timestamp:

```{r checkTimestamp}
results_df %>%
  filter(str_detect(results, '/data/cephfs/punim0010/data/Results')) %>%
  mutate(timestamp=basename(results)) %>%
  mutate(parsed=parse_date(timestamp)) %>%
  filter(is.na(parsed)) %>%
  kable()
```

Third pass. Make sure project name matches storage folder.

```{r checkProject}
results_df %>%
  mutate(folder=str_replace(results, '/data/cephfs/punim0010/data/Results/', '')) %>%
  separate(folder, 'store', '/', remove=FALSE, extra='drop') %>%
  filter(project != store) %>%
  kable()
```

